\documentclass[../../diss.tex]{subfiles}
\begin{document}

\chapter*{Part II.\newline Models of computation}

This part contains some preliminaries:
We formally define the models of computation on which the problems that we will study in the rest of this thesis are based.

\paragraph{Outline}

We start by introducing some basic notation.
Then, we define Turing machines and formulas in propositional logic and Presburger arithmetic; while they are not a main object of study in this thesis, they are needed for proving some complexity results.

In \cref{Chapter:LTSNFAs}, we define labeled transition systems as state-based language-generating mechanisms.
In particular, we focus on the class of finite automata.

\cref{Chapter:Grammars} is concerned with context-free grammars and higher-order recursion schemes as  grammar-based language-generating mechanisms.

Finally, \cref{Chapter:PetriNets} introduces Petri nets, a model for concurrent systems, and an extension thereof in the form of well-structured transition systems.

Some concepts that are only used in a single part of the thesis will be defined later: board games with perfect information in \cref{Chapter:Games}, fixed-point based techniques for systems of equations in \cref{Chapter:EDS}, and valence systems in \cref{Chapter:ValenceGames}.

\paragraph{Sources and publications}

This part contains almost no contributions by the author of the thesis; it merely presents commonly used models following standard textbooks.
References will be given in the main text.
There are two exceptions:
In \cref{Section:CFGOmega}, we present a novel way in which context-free grammars can be used to define languages of infinite words.
This is taken from the publication~\cite{MeyerMN17a}.
In \cref{Section:PNBPP}, we show that the word problem for BPP nets is $\NPTIME$-complete.
This result is not contained in the publication~\cite{AtigMMS17}, but its proof is an adaption of the proof of Theorem~24 from the full version~\cite{AtigMMS17a} of the paper.
We will discuss the contributions to these publications by the author of the thesis in more detail in \cref{Chapter:Contributions}.

\end{document}
