\documentclass[../../diss.tex]{subfiles}
\begin{document}

This chapter of the thesis is dedicated to studying the computation of the upward closures of Petri net languages.
The task is to compute an NFA whose language is the upward closure of the language of a given Petri net.
We first study this problem in the case of Petri net coverability languages, showing that the state complexity of the upward closure is doubly exponential, and that a corresponding NFA can be computed in doubly exponential time.
The proof techniques for the upper and lower bound use and extend the well-known results on coverability by \citea{Rackoff78} and~\citea{Lipton76}, respectively.

The doubly exponential complexity motivates studying restricted versions of the problem.
We do so in the second section by considering the upward closures of BPP net languages.
We show that in this case, both the state complexity and the time to compute a corresponding NFA are singly exponential.
% To prove the upper bound, we use a sort of pumping lemma.

Finally, we consider a different kind of restriction.
Instead of actually computing the upward closure, we consider the problem of checking whether a candidate SRE is included in it.
We will give more motivation for studying this problem in \cref{Section:PNSREUC}.
This restriction results in even better complexity: The problem is $\EXPSPACE$-complete for general Petri nets and $\NPTIME$-complete for BPP nets.
In both cases, the proofs make use of coverability having the same~complexity.

\end{document}
