\documentclass[../../diss.tex]{subfiles}
\begin{document}

We demonstrate the versatility of our technique for solving inclusion games based on effective denotational semantics by applying to it to games induced higher-order recursion schemes.
The setting is similar to the one in the previous chapter:
We consider a word-generating recursion scheme whose derivation processes are controlled by the players, and the goal of the game is membership \resp non-membership in a regular target language.
The approach that we take, however, is vastly different.

After giving the formal definition of HORS games, we consider the more general case of the system of equations associated to a deterministic HORS.\@
We design a model template that allows us to get a full model for interpreting such a system.
To instantiate the template, it is sufficient to provide a domain for kind ground and the interpretations for the nonterminals.
We apply this model template to the determinization of a game HORS and show that the associated least solution to the interpreted system of equations characterizes the winner of the game.
This leaves us with the problem that we cannot compute this least solution since we have used a domain that does not satisfy the ascending chain condition.

Returning to a more general setting, we present a framework for exact fixed-point transfer.
It allows us to transfer the properties of the least fixed point with respect to one model to the least fixed point with respect to another one.
We then use this framework to go from the aforementioned infinite domain to a model with a finite domain.
The latter allows us to compute the least solution while preserving the information about the winner of the game.

\paragraph{Publication}

The chapter presents material that has been published in the form of the paper~\cite{HagueMM17} (\resp its full version~\cite{HagueMM17a}).
Compared to the publication, the material has been improved.
We will discuss both the improvements and the authors author's contributions to the publication in \cref{Chapter:Contributions}.

\end{document}
